Co-methylation

### 2015-05-03 Rmarkdown/comethylation.Rmd
Edits to front matter and table of contents
Edit to co-methylation

\begin{chapabstract}
I describe two measures of within-sample co-methylation, \emph{within-fragment co-methylation} and \emph{correlation of $\beta$-values}. The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.

Using estimates of both within-fragment co-methylation and correlation of $\beta$-values, I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci. I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.

Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
\end{chapabstract}

@@

1. Define the pairs of methylation loci.
2. For each $IPD$, compute the correlation of $\beta$-values for all pairs of methylation loci separated by that $IPD$.

@@

I consider two different strategies for creating pairs of CpGs. Both strategies are based on CpGs in the reference genome after having filtered out those reference-specific CpGs. That is, I begin with $\mathcal{I}^{ref}$ and, based on the `Bis-SNP` output, I filter out loci that are not CpGs in the sample's genome. Rather than actually remove these sites, I simply set the corresponding $\beta$-value to `NA`. I also set the $\beta$-values of `missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.

@@

The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy. This is what is called an 'autocorrelation' analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$ by \citet{Cokus:2008fc}, although it is somewhat different to the traditional notion of an autocorrelation analysis. I refer to this as the $NIL \geq 0$ strategy. Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.

@@

Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two regions of the genome that have very different methylation dynamics. In other words, the vector of $\beta$-values for a given sample is highly non-stationary; the correlation between two loci depends on more than just the $IPD$. An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values. To do this, we can stratify our analysis by a genomic feature. While this does not eliminate this non-stationarity, it may help identify factors that drive some of the heterogeneity.

For any genomic feature, a pair of loci may be inside, outside or spanning the boundary of the feature. Figure \ref{fig:feature_status} illustrates the 'feature status' of some pairs of loci. Note that a pair where each locus is in a distinct feature from the same class is declared to be inside the feature, i.e. elements need not be in the same feature but rather in the same type of feature. The genomic feature should partition the genome so that each loci is either inside or outside of the feature. Since the number of 'spanning pairs' is substantially fewer than those inside or outside of the feature, and the boundaries of the feature are oftentimes fuzzy, I have excluded these 'spanning pairs' in the results shown below.

@@

Before computing correlations and trying to interpret these, it is a good idea to look at the data from which these correlations are computed. Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200, 2000$ with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample. It highlights two major problems.

@@

For all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets, I compute the Pearson and Spearman correlations using both the $NIL = 0$ and $NIL \geq 0$ strategies. I compute these genome-wide and after stratifying CpG pairs by whether they are in a CpG island. These analyses are performed separately for each strand and also for strand-collapsed $\beta$-values.

@@

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand. Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
%\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
%\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
%\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs %of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
%\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_5}
\end{figure}

\FloatBarrier
\subsubsection{Choice of correlation coefficient}

@@

\FloatBarrier
\subsection{$NIL = 0$ CpG pairs}

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} show the Spearman correlations of strand-collapsed $\beta$-values for $NIL = 0$ pairs of CpGs. Two things immediately stand out. Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$; pairs of CpGs separated larger distances are on average less correlated than pairs of CpGs separated by shorter distances. Secondly, the estimated correlations violently jump around for larger $IPD$s. The latter is simply due to the aforementioned lack of $NIL = 0$ pairs at larger $IPD$s with which to estimate these correlations. Therefore, it is the trend, rather than the individual values, that should be analysed in these plots.

The are some consistent patterns in comparing these trends across samples. The correlations very quickly drop well below $0.5$, within approximately $50$ bp for all the _EPISCOPE_ and _Seisenberger_ samples and most of the _Lister_ and _Ziller_ samples. Samples that are exceptions to this trend (_ADS_, _ADS-adipose_, _ADS-iPSC_, _FF_, _IMR90\_r1_, _IMR90\_r2_, _IMR90\_cell\_line_ and _HepG2\_cell\_line_) have a slower decay in these correlations and remain more highly correlated over $IPD = 2, \ldots, 1500$. These same samples are also notable because they have significant amounts of intermediate methylation (Section \ref{sec:genome-wide_distribution_of_beta_values}). I believe this is what drives the stronger and more slowly decaying correlations. The scatterplots of $\beta$-values for these samples will have more $\beta$-values in the middle of the scatterplot, and fewer in the corners, hence the stronger correlations.

Most samples have a correlation close to $1$ for neighbouring CpGs ($IPD = 2$). The notable exceptions are the embryonic stem cell replicates, _H1\_r1_ and _H1\_r2_ (Lister dataset), which begin with a maximum correlation of only around $0.5$. This is not observed in the other embryonic stem cells _H9_, _H9\_Laurent_, _HSF1_ and _J1\_1_), which makes it difficult to know whether $\beta$-values of embryonic stem cells are truly less correlated than other cell types.

The trend of the correlations for all these samples plateaus out close to zero, slightly higher for those samples with significant intermediate methylation. However, it should not be concluded from this that the methylation of CpGs separated by more than a few hundred base pairs are uncorrelated. Recall that these results are only for $NIL = 0$ pairs of CpGs, which are increasingly rare for larger $IPD$s. These results tell us about the 'first-order' correlations of $\beta$-values, which decay very quickly but whose importance also decreases quite quickly owing to the vast majority of CpGs being within $100$ bp of another CpG (e.g., Figure \ref{fig:IPD_ECDF_hg19} for humans).

@@

The above are genome-wide plots, all pairs of CpGs with a given $IPD$ are treated identically. However, there are good reasons to expect that the correlations of $\beta$-values, like the $\beta$-values themselves, might be influenced by their genomic context. To investigate this, we can stratify pairs of CpGs pairs by a genomic feature and repeat the analysis. The obvious feature to stratify on is CpG islands.

@@

The message from these plots is clear, $\beta$-values of $NIL = 0$ pairs within CpG islands are much more correlated than those pairs outside of islands. Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside islands is less influenced by $IPD$. In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD$. From these plots we can now see that the apparent sharp decline in correlations over the first $50$ bp from the genome-wide data is really driven by a shift from pairs that are frequently inside CpG islands to pairs that are mostly outside of islands.

@@

\FloatBarrier
\subsection{$NIL \geq 0$ CpG pairs}\label{sec:nil_geq_0_beta_cors}

@@

[^ipd_choice]: These correlations could be computed for $IPD > 2000$. However, constructing all $NIL \geq 0$ pairs requires much more time and memory than constructing all $NIL = 0$ pairs because there are many more such pairs. For this reason, and for compatability with the $NIL = 0$ results, I have focused on $IPD = 2, \ldots, 2000$ for the $NIL \geq 0$ pairs.

When we aggregate over as many factors as we are in these genome-wide $NIL \geq 0$ plots, we find that there is little difference in the correlation curves between samples. All samples have a very high correlation, close to $1$, for small $IPD$s that steadily decays to a lower correlation of $0.2-0.4$ by $IPD = 2000$.

@@

A limitation in to these analyses is they are based on correlations of measurements that are aggregates from many thousands of DNA molecules. We are not measuring co-methylation on the scale at which the biological process of DNA methylation acts, namely at the level of individual DNA molecules. As noted by \citet{Ball:2009kh}, "the clonal feature of Illumina sequencing [allows us] to investigate whether co-methylation occurs at the single-molecule level". We explore this topic in Section \ref{sec:wf_cometh}, but before doing so we first must first discuss the challenge of estimating meaningful measures of dependence in sparse $2 \times 2 \times K$ contingency tables. While perhaps not yet apparent, this will be essential for the analysis of within-fragment co-methylation.

@@

We will begin with an overview of the simplest case, estimating dependence in a $2 \times 2$ contingency table ($K = 1$). We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables. We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determining which is most appropriate for a study of within-fragment co-methylation.

@@

Consider two binary random variables, $X$ and $Y$. Let $\pi_{ij}$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$ ($i, j, = 1, 2$). We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$. The odds is a convenient summary of dependence in a $2 \times 2$ table.Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:

@@

Suppose we observe $(X, Y)$ for $n$ samples. Denote by $n_{ij} the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$. We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$. The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.

@@

The simplest estimator of $\theta$ is the unconditional maximum likelihood estimator, $\widehat{\theta_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$. The asymptotic standard error of $\widehat{\theta_{U}}$ is given by $\hat{\sigma}(\widehat{\theta_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. The asymptotic distribution of $\widehat{\theta_{U}}$ as $n \to \infty$ is $Gaussian(\theta, \hat{\sigma}(\theta)^{2})$.

A problem with the estimator $\widehat{\theta_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities. We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\theta_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\theta_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. \citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\theta_{U}}$ and $\widehat{\theta_{0.5}}$ have the same asymptotic distribution around $\theta$ as $n \to \infty$ but that $\widehat{\theta_{0.5}}$ has reduced first-order bias. The modified estimator, $\widehat{\theta_{0.5}}$, is therefore generally recommended.

Another alternative to $\widehat{\theta_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods. The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.

Estimation of the odds ratio is made more difficult when the contingency table is _sparse_. We say that a contingency table is sparse when some of the $n_{ij}$ are small. Sparsity can occur even when the sample size, $n$, is large. Sparsity becomes more of a problem as we move from a single $2 \times 2$ table to $K$ $2 \times 2$ tables or, equivalently, a $2 \times 2 \times K$ table.

@@

Suppose we now measure the same two binary variables $X$ and $Y$ in $K$ different strata. We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum. For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{+ 1k} = 1 - \pi_{+ 2k}, \pi_{1 + k} = 1 - \pi_{2 + k}$ to vary.

We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi}_{MH} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi}_{MH}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.

@@

1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
2. What factors influences this dependence?

>>>>

\begin{chapabstract}
This chapter describes two measures of within-sample co-methylation, \emph{within-fragment co-methylation} and \emph{correlation of $\beta$-values}. The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.

Using estimates of both within-fragment co-methylation and correlation of $\beta$-values, we explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci. We apply these methods to 40 whole-genome bisulfite-sequencing samples to study how co-methylation varies between different samples.

This chapter focuses on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
\end{chapabstract}

@@

1. Define the pairs of methylation loci.
2. For each intra-pair distance ($IPD$), compute the correlation of $\beta$-values for all pairs of methylation loci separated by that $IPD$.

@@

I consider two different strategies for creating pairs of CpGs. Both strategies are based on CpGs in the reference genome after having filtered out those reference-specific CpGs. That is, I begin with $\mathcal{I}^{ref}$ and, based on the `Bis-SNP` output, I filter out loci that are not CpGs in the sample's genome. Rather than actually remove these sites, I simply set the corresponding $\beta$-value to `NA`. I also set the $\beta$-values of 'missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.

@@

The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2$ to $2000$ bp, otherwise the number of pairs becomes unwieldy. This is what is called an 'autocorrelation' analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$ by \citet{Cokus:2008fc}, although it is somewhat different to the traditional notion of an autocorrelation analysis. I refer to this as the $NIL \geq 0$ strategy. Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.

@@

Two pairs of CpGs with an identical $IPD$ may come from two regions, even when using the $NIL = 0$ strategy. These regions mayhave very different methylation dynamics. In other words, the vector of $\beta$-values for a given sample is highly non-stationary; the correlation between two loci depends on more than just the $IPD$. An obvious variable to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values. To do this, we can stratify our analysis by a genomic feature. While this does not eliminate this heterogeneity, it may help identify factors that drive some of the variation.

For any genomic feature, a pair of loci may be inside, outside or spanning the boundary of the feature. Figure \ref{fig:feature_status} illustrates the 'feature status' of some pairs of loci. Note that a pair where each locus is in a distinct feature from the same class is declared to be inside the feature, i.e. elements need not be in the same feature but rather in the same type of feature. The genomic feature should partition the genome so that each locus is either inside or outside of the feature. Since the number of 'spanning pairs' is substantially fewer than those inside or outside of the feature, and the boundaries of the feature are oftentimes fuzzy, I have excluded these 'spanning pairs' in the results shown below.

@@

Before computing correlations and trying to interpret these, it is a good idea to look at the data from which these correlations are computed. Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200, 2000$ with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample from the _Lister_ dataset. It highlights two major problems.

@@

For all 40 samples in the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, I compute the Pearson and Spearman correlations using both the $NIL = 0$ and $NIL \geq 0$ strategies. I compute these genome-wide and after stratifying CpG pairs by whether they are in a CpG island. These analyses are performed separately for each strand and also for strand-collapsed $\beta$-values.

@@

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand. Since the trend in the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_5}
\end{figure}

@@

\FloatBarrier
\subsection{CpG pairs with $NIL = 0$}

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} show the Spearman correlations of strand-collapsed $\beta$-values for $NIL = 0$ pairs of CpGs. Two things immediately stand out. Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$; pairs of CpGs separated by larger distances are on average less correlated than pairs of CpGs separated by shorter distances. Secondly, the estimated correlations violently jump around for larger $IPD$s. The latter is simply due to the aforementioned lack of $NIL = 0$ pairs at larger $IPD$s with which to estimate these correlations. Therefore, it is the trend, rather than the individual values, that should be analysed in these plots.

There are some consistent patterns in comparing these trends across samples. The correlations very quickly drop well below $0.5$, within approximately $50$ bp for all the _EPISCOPE_ and _Seisenberger_ samples and most of the _Lister_ and _Ziller_ samples. Samples that are exceptions to this trend (_ADS_, _ADS-adipose_, _ADS-iPSC_, _FF_, _IMR90\_r1_, _IMR90\_r2_, _IMR90\_cell\_line_ and _HepG2\_cell\_line_) have a slower decay in these correlations and remain more highly correlated over $IPD = 2, \ldots, 1500$. These same samples are also notable because they have significant amounts of intermediate methylation (Section \ref{sec:genome-wide_distribution_of_beta_values}). I believe this is what drives the stronger and more slowly decaying correlations. The scatterplots of $\beta$-values for these samples will have more $\beta$-values in the middle of the scatterplot, and fewer in the corners, hence the stronger correlations.

Most samples have a correlation close to $1$ for neighbouring CpGs ($IPD = 2$). The notable exceptions are the embryonic stem cell replicates, _H1\_r1_ and _H1\_r2_ (Lister dataset), which begin with a maximum correlation of only around $0.5$. This is not observed in the other embryonic stem cells (_H9_, _H9\_Laurent_, _HSF1_ and _J1\_1_), which makes it difficult to know whether $\beta$-values of embryonic stem cells are truly less correlated than other cell types.

The trend of the correlations for all these samples plateaus out close to zero, slightly higher for those samples with significant intermediate methylation. It should not be concluded from this, however, that the methylation of CpGs separated by more than a few hundred base pairs are uncorrelated. Recall that these results are only for $NIL = 0$ pairs of CpGs, which are increasingly rare for larger $IPD$s. These results tell us about the 'first-order' correlations of $\beta$-values, which decay very quickly but whose importance also decreases quite quickly owing to the vast majority of CpGs being within $100$ bp of another CpG (e.g., Figure \ref{fig:IPD_ECDF_hg19} shows the distribution of $IPD$s for humans).

@@

Figures ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are genome-wide plots; all pairs of CpGs with a given $IPD$ are treated identically. However, there are good reasons to expect that the correlations of $\beta$-values, like the $\beta$-values themselves, might be influenced by their genomic context. To investigate this, we can stratify pairs of CpGs pairs by a genomic feature and repeat the analysis. The obvious feature to stratify on is CpG islands.

@@

The message from these plots is clear, $\beta$-values of $NIL = 0$ pairs within CpG islands are much more correlated than those pairs outside of islands. Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside CpG islands is less influenced by $IPD$. In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD$. From these plots we can now see that the apparent sharp decline in correlations over the first $50$ bp from the genome-wide data is really driven by a shift from pairs that are frequently inside CpG islands to pairs that are mostly outside of islands.

@@

\FloatBarrier
\subsection{CpG pairs with $NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}

@@

[^ipd_choice]: These correlations could be computed for $IPD > 2000$. However, constructing all $NIL \geq 0$ pairs requires much more time and memory than constructing all $NIL = 0$ pairs because there are many more such pairs. For this reason I have focused on $IPD = 2, \ldots, 2000$ for the $NIL \geq 0$ pairs.

When we aggregate over as many factors as we are in these genome-wide $NIL \geq 0$ plots, we find that there is little difference in the correlation curves between samples. All samples have a very high correlation, close to $1$, for small $IPD$s that steadily decays to a lower correlation of $0.2$ to $0.4$ by $IPD = 2000$.

@@

A limitation to these analyses is that they are based on correlations of measurements that are aggregates from many thousands of DNA molecules. We are not measuring co-methylation on the scale at which the biological process of DNA methylation acts, namely at the level of individual DNA molecules. As noted by \citet{Ball:2009kh}, "the clonal feature of Illumina sequencing [allows us] to investigate whether co-methylation occurs at the single-molecule level". We explore this topic in Section \ref{sec:wf_cometh}, but before doing so we first must first discuss the challenge of estimating meaningful measures of dependence in sparse $2 \times 2 \times K$ contingency tables. While perhaps not yet apparent, this will be essential for the analysis of within-fragment co-methylation.

@@

We will begin with an overview of the simplest case, estimating dependence in a $2 \times 2$ contingency table (i.e. $K = 1$). We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables. We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determinine which is most appropriate for a study of within-fragment co-methylation.

@@

Consider two binary random variables, $X$ and $Y$. Let $\pi_{ij}$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$ ($i, j, = 1, 2$). We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$. The odds ratio is a convenient summary of dependence in a $2 \times 2$ table. Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:

@@

Suppose we observe $(X, Y)$ for $n$ samples. Denote by $n_{ij}$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$. We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$. The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.

@@

The simplest estimator of $\theta$ is the unconditional maximum likelihood estimator, $\widehat{\theta}_{U} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$. The asymptotic standard error of $\widehat{\theta}_{U}$ is given by $\hat{\sigma}(\widehat{\theta}_{U}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. The asymptotic distribution of $\widehat{\theta}_{U}$ as $n \to \infty$ is $Gaussian(\theta, \hat{\sigma}(\theta)^{2})$.

A problem with the estimator $\widehat{\theta}_{U}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities. We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\theta}_{0.5} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $\hat{\sigma}(\widehat{\theta}_{0.5}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. \citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\theta}_{U}$ and $\widehat{\theta}_{0.5}$ have the same asymptotic distribution around $\theta$ as $n \to \infty$ but that $\widehat{\theta}_{0.5}$ has reduced first-order bias. The modified estimator, $\widehat{\theta}_{0.5}$, is therefore generally recommended.

Another alternative to $\widehat{\theta}_{U}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods. The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.

Estimation of the odds ratio is made more difficult when the contingency table is _sparse_. We say that a contingency table is sparse when some of the $n_{ij}$ are small. Sparsity can occur even when the sample size, $n$, is large. Sparsity becomes more of a problem as we move from a $2 \times 2$ table to a $2 \times 2 \times K$ table.

@@

Suppose we now measure the same two binary variables $X$ and $Y$ in $K$ different strata. We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum. We assume for now that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{+ 1k} = 1 - \pi_{+ 2k}, \pi_{1 + k} = 1 - \pi_{2 + k}$ to vary.

We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the odds ratio $\psi$ \citep[see][]{Breslow:1981ta}.

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi}_{MH} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi}_{MH}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the existance of two asymptotic forms of $2 \times 2 \times K$ contingency tables.

@@

1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
2. What factors influence this dependence?
